---
title: "Scaling phylogenies by phenotypic change"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling phylogenies by phenotypic change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(traittree)
```

## The idea

A time-scaled phylogeny measures every branch in the same currency: elapsed
time. That is often not the currency a question is asked in. Two lineages that
split ten million years ago may have diverged enormously or barely at all, and a
diversity metric built on time cannot tell those cases apart.

This package rewrites branch lengths so that they measure phenotypic change
instead. Ancestral trait values are reconstructed at every node, and each branch
becomes the distance in trait space between the nodes at its ends.

The package ships simulated data for 60 imaginary species: a phylogeny, three
traits, and occurrences over a 10 by 10 grid.

```{r data}
example_tree

head(example_traits)

head(example_occurrences)
```

## Scaling a tree by trait change

```{r scale}
trait_tree <- scale_branches_multidimensional(tree = example_tree,
                                              traits = example_traits)
```

The topology is untouched; only the branch lengths change.

```{r compare, fig.alt = "Trait change against elapsed time for each branch"}
plot(example_tree$edge.length, trait_tree$edge.length,
     xlab = "branch length (time)",
     ylab = "branch length (trait change)",
     pch = 16, col = "#4477aa")
```

The relationship is positive — more time generally means more change — but far
from tight, and that scatter is the point of the exercise.

Passing `rate = TRUE` divides by the original branch lengths, giving rates of
change rather than absolute amounts:

```{r rate}
rate_tree <- scale_branches_multidimensional(tree = example_tree,
                                             traits = example_traits,
                                             rate = TRUE)
```

## Traits have units, and units should not matter

Branch lengths here are Euclidean distances across traits. Body mass in grams
runs to hundreds; bill depth in millimetres runs to tens. Left alone, the first
trait would dominate simply because its numbers are bigger, and re-expressing it
in kilograms would change the answer.

By default each trait is therefore divided by its standard deviation across
species. Compare the two:

```{r scaling}
scaled <- scale_branches_multidimensional(example_tree, example_traits)

raw <- suppressWarnings(
  scale_branches_multidimensional(example_tree, example_traits,
                                  scale_traits = FALSE))

cor(scaled$edge.length, raw$edge.length)
```

Using the traits as supplied emits a warning naming the traits whose spreads
differ most, because that choice is rarely deliberate. It is sometimes the right
one: if your traits are already on a common scale — all log-transformed, say —
their relative spread is itself meaningful, and scaling would flatten a real
difference in evolutionary rate. That is what `scale_traits = FALSE` is for.

`weights` gives finer control, applied to the squared differences:

```{r weights}
mass_heavy <- scale_branches_multidimensional(example_tree, example_traits,
                                              weights = c(2, 1, 1))
```

## Carrying uncertainty through

Ancestral states are estimates. Treating them as known throws away everything we
know about how uncertain they are.

`scale_branches_multidimensional_with_variation()` draws trees from the joint
posterior of the ancestral states. Each tree is one coherent realisation: a node
takes a single value shared by every branch meeting there, and the correlations
between neighbouring nodes and between traits are preserved.

```{r posterior}
trees <- scale_branches_multidimensional_with_variation(tree = example_tree,
                                                        traits = example_traits,
                                                        n_trees = 50)

class(trees)
```

The spread across trees is the reconstruction uncertainty:

```{r spread, fig.alt = "Uncertainty in each branch length against its estimated length"}
branch_lengths <- vapply(trees, function(x) x$edge.length,
                         numeric(nrow(example_tree$edge)))

plot(rowMeans(branch_lengths), apply(branch_lengths, 1, sd),
     xlab = "mean branch length", ylab = "standard deviation",
     pch = 16, col = "#4477aa")
```

Why the joint posterior rather than each node's own variance? Because branch
lengths are *differences* between adjacent nodes, and adjacent nodes are
strongly correlated. Sampling each node independently ignores that correlation
and inflates branch lengths, worst on the short branches where trait-scaled
trees carry the most information.

## Within-species variation

`Rphylopars::phylopars()` estimates a within-species error term whenever the
argument is left unset. With a single observation per species there is nothing
to estimate it from, so this package decides from the data: the term is used
only when at least one species has two or more observations of a trait.

```{r pheno, eval = FALSE}
# forced on or off, whatever the data look like
scale_branches_multidimensional(example_tree, example_traits, pheno_error = TRUE)
```

Replicate observations are simply extra rows for the same species. When
replication is very sparse — a single duplicated row in a large table, which is
more often a data-entry artifact than a measurement — you get a warning, because
that one row would otherwise switch the model silently.

## From trees to maps

`data_matching()` prunes a tree and an occurrence table down to their shared
species, and `richness_raster()` counts species per cell:

```{r spatial}
template <- raster::raster(nrows = 10, ncols = 10)

matched <- data_matching(phylogeny = example_tree,
                         occurrences = example_occurrences)

richness <- richness_raster(template.raster = template,
                            occurrences = example_occurrences)
```

`calculate_pd_metric_rasters()` puts the whole workflow together, returning
phylogenetic diversity computed on both the time-scaled and trait-scaled trees:

```{r pd}
result <- calculate_pd_metric_rasters(occurrences = example_occurrences,
                                      phylogeny = example_tree,
                                      traits = example_traits,
                                      template.raster = template)

names(result$pd_stack)
```

Passing `n_trees` propagates the reconstruction uncertainty all the way to the
maps. The time-based layers are unaffected — they do not involve the traits — but
the trait-based ones are replaced by a mean, a standard deviation, and an
interval across trees:

```{r pd_uncertainty}
sampled <- calculate_pd_metric_rasters(occurrences = example_occurrences,
                                       phylogeny = example_tree,
                                       traits = example_traits,
                                       template.raster = template,
                                       n_trees = 25)

names(sampled$pd_stack)

summary(raster::values(sampled$pd_stack[["pd_traits_sd"]]))
```

That last column is the part a single point estimate cannot show you: how much
of the spatial pattern in trait-based diversity is real, and how much is an
artifact of not knowing the ancestral states.

If you already have trait-scaled trees in hand, pass them directly with
`trait_phylo` instead of having them rebuilt.

## Single traits, one tree at a time

Two older functions handle the univariate case. `scale_branches_by_traits_rphylopars()`
returns one tree per trait:

```{r univariate}
per_trait <- scale_branches_by_traits_rphylopars(tree = example_tree,
                                                 traits = example_traits)

names(per_trait)
```

and `scale_branches_by_traits_fastAnc()` takes a single trait as a named vector,
reconstructing it with `phytools::fastAnc()`:

```{r fastanc}
body_mass <- example_traits$body_mass_g
names(body_mass) <- example_traits$species

mass_tree <- scale_branches_by_traits_fastAnc(tree = example_tree,
                                              traits = body_mass)
```

Both accept `percent = TRUE`, which expresses change relative to the ancestral
value at the start of each branch. That measure assumes a denominator safely
away from zero and positive; log-transformed or centred traits satisfy neither,
and you will be warned when they do not.
