---
title: "Scaling phylogenies by phenotypic change"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling phylogenies by phenotypic change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(traittree)
```

## The idea

A time-scaled phylogeny measures every branch in the same currency: elapsed
time. That is often not the currency a question is asked in. Two lineages that
split ten million years ago may have diverged enormously or barely at all, and a
diversity metric built on time cannot tell those cases apart.

This package rewrites branch lengths so that they measure phenotypic change
instead. Ancestral trait values are reconstructed at every node, and each branch
becomes the distance in trait space between the nodes at its ends.

The package ships simulated data for 60 imaginary species: a phylogeny, three
traits, and occurrences over a 10 by 10 grid.

```{r data}
example_tree

head(example_traits)

head(example_occurrences)
```

## Scaling a tree by trait change

```{r scale}
trait_tree <- scale_branches_multidimensional(tree = example_tree,
                                              traits = example_traits)
```

The topology is untouched; only the branch lengths change.

```{r compare, fig.alt = "Trait change against elapsed time for each branch"}
plot(example_tree$edge.length, trait_tree$edge.length,
     xlab = "branch length (time)",
     ylab = "branch length (trait change)",
     pch = 16, col = "#4477aa")
```

The relationship is positive (more time generally means more change) but far
from tight, and that scatter is the point of the exercise.

Passing `rate = TRUE` divides by the original branch lengths, giving rates of
change rather than absolute amounts:

```{r rate}
rate_tree <- scale_branches_multidimensional(tree = example_tree,
                                             traits = example_traits,
                                             rate = TRUE)
```

## Traits have units, and units should not matter

Branch lengths here are Euclidean distances across traits. Body mass in grams
runs to hundreds; bill depth in millimetres runs to tens. Left alone, the first
trait would dominate simply because its numbers are bigger, and re-expressing it
in kilograms would change the answer.

By default each trait is therefore divided by its standard deviation across
species. Compare the two:

```{r scaling}
scaled <- scale_branches_multidimensional(example_tree, example_traits)

raw <- suppressWarnings(
  scale_branches_multidimensional(example_tree, example_traits,
                                  scale_traits = FALSE))

cor(scaled$edge.length, raw$edge.length)
```

Using the traits as supplied emits a warning naming the traits whose spreads
differ most, because that choice is rarely deliberate. It is sometimes the right
one: if your traits are already on a common scale, all log-transformed, say,
their relative spread is itself meaningful, and scaling would flatten a real
difference in evolutionary rate. That is what `scale_traits = FALSE` is for.

`weights` gives finer control, applied to the squared differences:

```{r weights}
mass_heavy <- scale_branches_multidimensional(example_tree, example_traits,
                                              weights = c(2, 1, 1))
```

## Carrying uncertainty through

Ancestral states are estimates. Treating them as known throws away everything we
know about how uncertain they are.

`scale_branches_multidimensional_with_variation()` draws trees from the joint
posterior of the ancestral states. Each tree is one coherent realisation: a node
takes a single value shared by every branch meeting there, and the correlations
between neighbouring nodes and between traits are preserved.

```{r posterior}
trees <- scale_branches_multidimensional_with_variation(tree = example_tree,
                                                        traits = example_traits,
                                                        n_trees = 50)

class(trees)
```

The spread across trees is the reconstruction uncertainty:

```{r spread, fig.alt = "Uncertainty in each branch length against its estimated length"}
branch_lengths <- vapply(trees, function(x) x$edge.length,
                         numeric(nrow(example_tree$edge)))

plot(rowMeans(branch_lengths), apply(branch_lengths, 1, sd),
     xlab = "mean branch length", ylab = "standard deviation",
     pch = 16, col = "#4477aa")
```

Why the joint posterior rather than each node's own variance? Because branch
lengths are *differences* between adjacent nodes, and adjacent nodes are
strongly correlated. Sampling each node independently ignores that correlation
and inflates branch lengths, worst on the short branches where trait-scaled
trees carry the most information.

## Within-species variation

`Rphylopars::phylopars()` estimates a within-species error term whenever the
argument is left unset. With a single observation per species there is nothing
to estimate it from, so this package decides from the data: the term is used
only when at least one species has two or more observations of a trait.

```{r pheno, eval = FALSE}
# forced on or off, whatever the data look like
scale_branches_multidimensional(example_tree, example_traits, pheno_error = TRUE)
```

Replicate observations are simply extra rows for the same species. When
replication is very sparse (a single duplicated row in a large table, which is
more often a data-entry artifact than a measurement) you get a warning, because
that one row would otherwise switch the model silently.

## From trees to maps

`data_matching()` prunes a tree and an occurrence table down to their shared
species, and `richness_raster()` counts species per cell:

```{r spatial}
template <- raster::raster(nrows = 10, ncols = 10)

matched <- data_matching(phylogeny = example_tree,
                         occurrences = example_occurrences)

richness <- richness_raster(template.raster = template,
                            occurrences = example_occurrences)
```

`calculate_pd_metric_rasters()` puts the whole workflow together, returning
phylogenetic diversity computed on both the time-scaled and trait-scaled trees:

```{r pd}
result <- calculate_pd_metric_rasters(occurrences = example_occurrences,
                                      phylogeny = example_tree,
                                      traits = example_traits,
                                      template.raster = template)

names(result$pd_stack)
```

Passing `n_trees` propagates the reconstruction uncertainty all the way to the
maps. The time-based layers are unaffected, since they do not involve the
traits, but the trait-based ones are replaced by a mean, a standard deviation,
and an interval across trees, whose bounds `probs` controls:

```{r pd_uncertainty}
sampled <- calculate_pd_metric_rasters(occurrences = example_occurrences,
                                       phylogeny = example_tree,
                                       traits = example_traits,
                                       template.raster = template,
                                       n_trees = 25)

names(sampled$pd_stack)

summary(raster::values(sampled$pd_stack[["pd_traits_sd"]]))
```

That last column is the part a single point estimate cannot show you: how much
of the spatial pattern in trait-based diversity is real, and how much is an
artifact of not knowing the ancestral states.

If you already have trait-scaled trees in hand, pass them directly with
`trait_phylo` instead of having them rebuilt, and set `verbose = TRUE` to watch
the loop over cells.

## The other route: blending distances

Rescaling branch lengths is one way to combine trait and phylogenetic
information. The other is to leave the tree alone and mix the two distance
matrices after the fact, following the functional-phylogenetic distance of
Cadotte, Albert and Walker (2013):

$$FPD = (a P^p + (1 - a) F^p)^{1/p}$$

where $P$ and $F$ are the phylogenetic and trait distances. Each is divided by
its maximum first, so that $a$ weights the two sources against each other rather
than against whatever units they happen to be measured in.

```{r blend}
trait_matrix <- scale(as.matrix(example_traits[, -1]))
row.names(trait_matrix) <- example_traits$species

blended <- blend_distances(phylo_dist = example_tree,
                           trait_dist = trait_matrix,
                           a = 0.25)

round(blended[1:4, 1:4], 2)
```

The traits are scaled here for the reason given above: a Euclidean distance
taken across grams and millimetres belongs mostly to whichever trait has the
bigger numbers. `blend_distances()` will not do it for you, because it accepts a
ready-made distance matrix just as readily as a trait table, and cannot tell
which it has been handed. Species are matched by name rather than by position,
so the two inputs need not be in the same order, and the endpoints recover the
inputs: `a = 1` is the scaled phylogenetic distance and `a = 0` the scaled trait
distance.

### Choosing the weight

There is no default for `a`, and no established estimator in the literature,
where the convention is to compute the blend across the whole range and read the
curve as a diagnostic. `tune_blend_weight()` chooses one value from the data
instead. It holds out each measured trait in turn and asks which weight best
reproduces that trait's distance structure from the remaining traits and the
phylogeny, on the reasoning that a held-out measured trait stands in for the
trait axes that were never measured, which are what the blend is trying to
describe in the first place. At least two traits are needed: with one, holding
it out leaves no trait component and `a` has no effect.

```{r tune}
fit <- tune_blend_weight(phylo_dist = example_tree, traits = trait_matrix)

fit$a

fit$a_plateau
```

```{r tune_curve, fig.alt = "Mean cross-validated score against the weight on the phylogeny"}
plot(fit$a_grid, fit$mean_score, type = "b", pch = 16, col = "#4477aa",
     xlab = "weight on the phylogeny (a)", ylab = "mean fold score")

abline(v = fit$a_plateau, lty = 2, col = "grey50")
```

The traits in the example data were evolved on the example tree, so the
phylogeny predicts them well and the weight comes back high. The shape of the
curve matters more than its peak: it climbs steeply out of `a = 0` and then goes
flat, and the dashed lines mark `a_plateau`, the range of weights whose scores
sit within one standard error of the best. Here that range covers almost the
whole grid, which is the honest summary. On real data the single best weight sat
at 0.125 to 0.25 for mammals and birds, with a curve just as flat across that
range, and the argmax was biased low. Read the plateau alongside the weight, and
treat a curve that is flat everywhere as the message it is: nothing in the data
distinguishes the candidates, because the measured traits carry no signal about
one another and the tree carries none about them.

`p` is fixed at 2, the convention in Cadotte et al. (2013), and is deliberately
not tuned. Scoring a grid over both parameters showed the achievable performance
moving by only 0.04 to 0.07 in correlation across `p` from 0.1 to 5,
cross-validation recovering the best `p` in 9% to 55% of subsets depending on
the grid, and tuning `p` alongside `a` beating a fixed `p = 2` in about a fifth
of them. The weight is worth estimating; the norm is not. Values below 1 are
accepted but warn, since they break the triangle inequality and so return a
dissimilarity rather than a metric.

### Which route to use

The objective decides, more than the data do. When the goal is to accumulate
diversity over a set of species (complementarity, surrogacy, reserve selection),
a trait-scaled tree did better than a blend in our simulations and on real
mammal and bird data. When the goal is to rank individual species by how
distinctive they are (the EDGE and FUSE style of question), the ordering
reverses and a blend did better, up to about 75% trait coverage, above which the
measured traits alone are enough.

The two routes also differ in what they return. Blending gives a distance
matrix, so Faith's PD and fair-proportion evolutionary distinctiveness are not
defined on its output and `calculate_pd_metric_rasters()` cannot take it, while
rescaling returns a phylogeny with the original topology. Rescaling is linear in
tree size and has no parameter to choose; blending needs an $n$ by $n$ matrix
and a weight. Against all that, a blend uses the measured trait distances
exactly, where rescaling passes them through an ancestral reconstruction that
shrinks them.

Nothing stops the phylogenetic component from being a trait-scaled tree:

```{r blend_component}
alt <- blend_distances(phylo_dist = trait_tree,
                       trait_dist = trait_matrix,
                       a = 0.25)
```

That helped in simulation. It did not help on real mammal data, where the dated
tree was the equal or better component at every trait count tried. What a
phylogenetic component contributes to a blend appears to be its independence
from the measured traits, which rescaling gives up. Worth trying, in other
words, but not worth assuming.

## Single traits, one tree at a time

Two older functions handle the univariate case. `scale_branches_by_traits_rphylopars()`
returns one tree per trait:

```{r univariate}
per_trait <- scale_branches_by_traits_rphylopars(tree = example_tree,
                                                 traits = example_traits)

names(per_trait)
```

and `scale_branches_by_traits_fastAnc()` takes a single trait as a named vector,
reconstructing it with `phytools::fastAnc()`:

```{r fastanc}
body_mass <- example_traits$body_mass_g
names(body_mass) <- example_traits$species

mass_tree <- scale_branches_by_traits_fastAnc(tree = example_tree,
                                              traits = body_mass)
```

Both accept `percent = TRUE`, which expresses change relative to the ancestral
value at the start of each branch. That measure assumes a denominator safely
away from zero and positive; log-transformed or centred traits satisfy neither,
and you will be warned when they do not.
